A general model of microbial interactions in a chemostat

### 2019-11-28 tex/Tracking.tex
changes in tex

\begin{abstract}
	This article deals with the inclusion of microbial ecology in bioprocess modelling. The first part exposes the mathematical analysis of a model that may be framed within the class of Lotka-Volterra models fitted to experimental data in a chemostat setting. The limitations and the insights of such an approach are discussed. The main novelty of this work lies in the second part, where the use of an optimal tracking technique for the integration of genetic sequencing in chemostat models is presented. The method revisits the data used for the first model presented and proposes a way to calibrate and discriminate . 
\end{abstract}

@@

\subsection{Stoichiometric Equations}
A cascade (bio)reaction process is considered. Suppose $n$ different OTU are present in the chemostat. The cascade reaction refers to the situation where a group of microorganisms ($G_1 \subset [n] $) consumes a substrate $s_1$ and produces $s_2$ and biomass, while another group of microorganisms ($G_2\subset [n]$) consumes $s_2$ and produces $s_3$ and biomass. $G_1$ and $G_2$ are called functional groups. The number of organisms in each functional will be denoted $n_1$ and $n_2$ respectively, that is $\vert G_1 \vert = n_1$ and $\vert G_2 \vert = n_2$ . This work treats the case when the $G_1$ and $G_2$ are disjoint sets:

@@

The terms $y_i$ are known as yields, it represents the number of grams of biomass produced per gram of substrate consumed. In this work the unit grams of dry biomass per gram of substrate is used for yields. Furthermore, for each $ i \in [n]$, OTU $i$ is characterized by its process rate (also known as growth function) $\mu_i(s,x)$. Notice that for being as generic as possible, the growth rate may be a function of the whole state in order to model the influence of all OTU on the growth rates of others.

\begin{align}
\label{Reaction1} \tag{R G1} 	s_1 \stackrel{\mu_i(s,x)}{\longrightarrow} s_2 + y_ix_1 \quad \forall i \in G_1 \\
\label{Reaction2} \tag{R G2} s_2\stackrel{\mu_i(s,x)}{\longrightarrow} s_3 + y_ix_i \quad \forall i \in G_2
\end{align}

\subsection{Kinetic Equations}
The growth rate of each species is given by:
\begin{align}
\mu_i(s,x) &= \bar{\mu}_i \dfrac{s_1}{K_i + s_1}\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right) \quad \forall i \in G_1 \\
\mu_i(s,x) &= \bar{\mu}_i \dfrac{s_2}{K_i + s_2}\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right) \quad \forall i \in G_2
\end{align}
The term $\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right)$ accounts for pairwise interactions affecting the growth rate of each OTU. Let $A$ denote the matrix
with entries $a_{ij}$, which will be referred to as the interaction matrix.

\subsection{Mass Balance Equations}

Consider the scenario of an continuous and homogeneous reactor: the input flow is the same as the output flow, with a dilution rate of $D$. The input flow contains a concentration $s_{in}$ of substrate $s_1$. Note that $s_{in}$ and $D$ are time dependent. Each OTU grows at a rate $\mu_i(s,x)$. System \eqref{system} represents this situation. A specific case of $\mu_i(s,x)$ is given by \cite{Dumont2016}, and the first part of this article deals with the mathematical analysis of that model. A second part then generalizes its growth function and calibrates it against experimental data using an optimal control approach.

\begin{align} 
\label{system}
\begin{array}{cl}
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_1\\
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_2 \\
\dot{s_1} =& \displaystyle (s_{in}-s_1)D-\sum\limits_{i\in G_1} \frac{1}{y_i}\mu_i(s_1,x) x_i  \\
\dot{s_2} = & \displaystyle -s_2D+\sum\limits_{i\in G_1}\frac{1}{y_i}\mu_i(s,x)x_i 		-\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i  \\
\dot{s_3} =&  \displaystyle -s_3D+\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i 
\end{array}
\end{align}

@@

\begin{theo}
	\label{theoWellPosedness}
	For initial conditions $(x_1(0),\dots,x_n(0),s_1(0),s_2(0),s_3(0)))\in \Omega$, there exists a constant $M>0$ such that for every matrix $A$ satisfying $\Vert A \Vert_{\infty} \leq M $, the solutions remain in $\Omega$ and are bounded.
\end{theo}

@@

In this section analytical expressions for equilibrium points are derived. Analytic expressions of equilibrium points are obtained, however no analytic expression concerning the stability of such points is presented. To answer the question of stability a numeric scheme is used by evaluating the Jacobian at the equilibrium point. At the end of the section an algorithm is provided for exploring all the possible equilibria.

@@

The former function places a vector $v$ in the diagonal of a matrix. By noting $A$ the matrix with coefficients $a_{ij}$, system \eqref{system} is restated as follows:
\begin{align}
\label{Eq1}	\dot{x} = & \diag(x)(\diag(f(s))(1_{n\times 1} + Ax) - D_{n\times 1}) \\
\label{Eq2}	\dot{s_1} = & (s_{in}-s_1)D + \begin{bmatrix}
-(1_{n_1\times 1} +A_1x) \\0
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\
\label{Eq3}	\dot{s_2} = & -s_2D+
\begin{bmatrix}
1_{n_1\times 1} +A_1x \\ -(1_{n_2\times 1} +A_2x)
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\	
\label{Eq4}	\dot{s_3} = & -s_3D+\begin{bmatrix}
0	\\(1_{n_2\times 1} +A_2x) 
\end{bmatrix}^\top \diag(k) \diag(f(s))x 
\end{align}

%
%The following notation was used:
%$A = \begin{bmatrix}
%A_1\\A_2
%\end{bmatrix}$.
%
%Where $A_1$ is a matrix of dimension $n_1 \times n$ corresponding to the interaction coefficients of OTU $i \in G_1$. $A_2$ is a matrix of dimension $n_2 \times n$ rows corresponding to the interaction coefficients of OTU $i \in G_2$.

@@

Note $J_I(x)$ the Jacobian matrix of function $\mathcal{I}$. then a first order approximation of $\I(\cdot)$ centred at $\bar{x}$ gives: $\I(x) = \I(\bar{x}) + J_I(\bar{x})(x-\bar{x}) + o(\Vert x- \bar{x} \Vert)$. I $\bar{x}= 0$, one recovers the growth expression from the previous section and $J_I(\bar{0})$ can be seen as the interaction matrix from model \eqref{system}.

@@

The control $u(t)$ is intended to drive the system to be near a desired output $z(t)$, which in this context are the measurements of the concentrations of OTU and metabolites. The term $\lambda \Vert(u -\vec{1}) \Vert_R$, was added for two reasons:

@@

The problem of approximating the solution of the system to a desired reference ($z$ in this case) is called the optimal tracking problem. For solving such a problem the approach developed by Cimen \textit{et al.} \cite{Cimen2004} was adapted to our problem. The method proposed involves the resolution of Approximating Sequences of Ricatti Equations (ASRE).

@@

Two points of departure from the original method are made clear, details can be seen in the appendix: first the tracking problem does not consider a constrained control, so one should consider a reformulation as proposed by the same author in other works \cite{Cimen2008}. Nevertheless, the methods were directly used, and an explicit constraint in the synthesis of the control was added for the control to remain within the desired bounds.  Even though this is probably suboptimal it worked in a first implementation.

Second in their problem they used an approximation of their original dynamics for calculating the trajectories and subsequently proved that such linearisation converged to the original dynamics. In the case here presented the linearisation of the dynamics was not necessary.

@@

Finally for the parametric identification of a particular class of interaction functions the following static optimization problem is proposed. Suppose $u(t)$ is the optimal control of problem \eqref{Optimal Control}, and $x(t)$ is the associated trajectory. Suppose a set of parameters $\theta \in \Theta$, such that the interaction function has a explicit form that depends on this parameters $I_{\theta}(x)$ ($\theta$ would represent matrix $A$ from section 2 for example). Then one solves the following optimization problem:
\begin{align}
\label{Parameter Identification} \min & \quad \Vert I_{\theta}(x(\cdot)) - u(\cdot) \Vert_{L^2([0,T])} \\
s.t. &  \quad \theta \in \Theta \notag \\
& \quad t \in [0,T] \notag
\end{align}

@@

For compensating the lack of a theoretical optimal control, the approach presented in here was first tested against synthetic data. The test consisted in simulating system \eqref{Controlsystem} with $I(x) = 1 + Ax$, (with $A$ in figure \ref{Comparison Interaction Matrices}) for $n_1 = 10$ and $n_2 = 5$. In all figures, dotted lines represent the synthetic data, while the continuous lines represent the method's output.

The choice of matrices $Q$ and $R$ are always of discussion in a quadratic regulator. $Q$ represents the covariance matrix of measurements, while the interpretation of $R$ is not clear. In the simulations here presented $Q = I_n$ and $R = \begin{bmatrix}
10^{-4}I_{n_1} & 0 \\ 0 & 10^{-5}I_{n_2}
\end{bmatrix}$, found by trial and error in order for the control to fit accurately the species abundance. The rational behind choosing a smaller value for the group $G_2$ is related to the one order of magnitude difference in yields between $G_1$ and $G_2$.

@@

The obtained matrix from the fit can be seen an compared to the original matrix for simulating data in \eqref{Comparison Interaction Matrices}. It can be seen that the new matrix is very different from the original one, in both order of magnitude and sign. Implying that the model is probably not identifiable. The system was simulated again with the new matrix, and one can see that it in general terms it captures the outcome of the system as shown in figures \ref{AOB LV simulation}, \ref{NOB LV simulation}, and \ref{Metabolites LV simulation}. The final message of this section is that if the interaction function is of the form $1+Ax$, one could be able to get a fit from the data with the method here presented.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_AOB_Iter_5_LV}
	\caption{AOB trajectories using the identified matrix seen in \eqref{Comparison Interaction Matrices}.}
	\label{AOB LV simulation}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_NOB_Iter_5_LV}
	\caption{NOB trajectories using the identified matrix seen in \eqref{Comparison Interaction Matrices}.}
	\label{NOB LV simulation}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_metabolites_Iter_5_LV}
	\caption{Metabolites trajectories using the identified matrix seen in \eqref{Comparison Interaction Matrices}.}
	\label{Metabolites LV simulation}
\end{figure}

>>>>

\begin{abstract}
	This article deals with the inclusion of microbial ecology in bioprocess modelling. The first part presents the mathematical analysis of a model that may be framed within the class of Lotka-Volterra models fitted to experimental data in a chemostat setting. The limitations and the insights of such an approach are discussed. The main novelty of this work lies in the second part, where the use of an optimal tracking technique for the integration of genetic sequencing in chemostat models is presented. The method revisits the data used for the first model presented and proposes a way to calibrate and discriminate for a class . 
\end{abstract}

@@

\subsection{Stoichiometric Equations}
A cascade (bio)reaction process is considered. Suppose $n$ different OTU are present in the chemostat. The cascade reaction refers to the situation where a group of microorganisms ($G_1 \subset [n] $) consumes a substrate $s_1$ and produces $s_2$ and biomass, while another group of microorganisms ($G_2\subset [n]$) consumes $s_2$ and produces $s_3$ and biomass. $G_1$ and $G_2$ are called functional groups. The number of organisms in each functional will be denoted $n_1$ and $n_2$ respectively, that is $\vert G_1 \vert = n_1$ and $\vert G_2 \vert = n_2$ . This work treats the case when $G_1$ and $G_2$ are disjoint sets:

@@

The terms $y_i$ are known as yields, it represents the number of grams of biomass produced per gram of substrate consumed. In this work the unit grams of dry biomass per gram of substrate is used for yields. Furthermore, for each $ i \in [n]$, OTU $i$ is characterized by its process rate (also known as growth function) $\mu_i(s,x)$. Notice that for being as generic as possible, the growth rate may be a function of the whole state in order to model the influence of all OTU on the growth rates of others.
\begin{align}
\label{Reaction1} \tag{R G1} 	s_1 \stackrel{\mu_i(s,x)}{\longrightarrow} s_2 + y_ix_1 \quad \forall i \in G_1 \\
\label{Reaction2} \tag{R G2} s_2\stackrel{\mu_i(s,x)}{\longrightarrow} s_3 + y_ix_i \quad \forall i \in G_2
\end{align}

An example of this process is the nitrification process where group $G_1$ is known as Ammonium oxidizer Bacteria (AOB), and group $G_2$ is known as Nitrite oxidizing Bacteria (NOB) \cite{SHARMA1977897}.
\subsection{Mass Balance Equations}

Consider the scenario of an continuous and homogeneous reactor: the input flow is the same as the output flow, with a dilution rate of $D$. The input flow contains a concentration $s_{in}$ of substrate $s_1$. Note that $s_{in}$ and $D$ are time dependent. Each OTU grows at a rate $\mu_i(s,x)$. System \eqref{system} represents this situation. A specific case of $\mu_i(s,x)$ is given in the next subsection.

\begin{align} 
\label{system}
\begin{array}{cl}
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_1\\
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_2 \\
\dot{s_1} =& \displaystyle (s_{in}-s_1)D-\sum\limits_{i\in G_1} \frac{1}{y_i}\mu_i(s,x) x_i  \\
\dot{s_2} = & \displaystyle -s_2D+\sum\limits_{i\in G_1}\frac{1}{y_i}\mu_i(s,x)x_i 		-\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i  \\
\dot{s_3} =&  \displaystyle -s_3D+\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i 
\end{array}
\end{align}

\subsection{Kinetic Equations}
In the work of Dumont \textit{et al.} \cite{Dumont2016} the growth rate seen in equations \eqref{gLV growth1} and \eqref{gLV growth2} was calibrated against experimental data for the two most abundant OTU of each functional group.

\begin{align}
\label{gLV growth1}\mu_i(s,x) &= \bar{\mu}_i \dfrac{s_1}{K_i + s_1}\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right) \quad \forall i \in G_1 \\
\label{gLV growth2}\mu_i(s,x) &= \bar{\mu}_i \dfrac{s_2}{K_i + s_2}\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right) \quad \forall i \in G_2 
\end{align}

The term $\left(1+\sum \limits_{j\in [n]} a_{ij} x_{j} \right)$ accounts for pairwise interactions affecting the growth rate of each OTU. Let $A$ denote the matrix
with entries $a_{ij}$, which will be referred to as the interaction matrix. The following section of this article deals with the mathematical analysis of model \eqref{system} with growth rates given by \eqref{gLV growth1} and \eqref{gLV growth2}.

@@

\begin{theo}
	\label{theoWellPosedness}
	For initial conditions $(x_1(0),\dots,x_n(0),s_1(0),s_2(0),s_3(0)))\in \Omega$, there exists a constant $M>0$ such that for every matrix $A$ satisfying $\Vert A \Vert_{\infty} \leq M $, the solutions of system \eqref{system} with growth rates given by \eqref{gLV growth1} and \eqref{gLV growth2}, remain in $\Omega$ and are bounded.
\end{theo}

@@

In this section analytical expressions for equilibrium points are derived. However no analytic expression concerning the stability of such points is presented. To answer the question of stability a numeric scheme is used by evaluating the Jacobian at the equilibrium point. At the end of the section an algorithm is provided for exploring all the possible equilibria.

@@

The former function places a vector $v$ in the diagonal of a matrix. Recalling $A$ the matrix with coefficients $a_{ij}$. The following notation is introduced.

\begin{align} A = \begin{bmatrix}
A_1\\A_2
\end{bmatrix}\end{align}
 Where $A_1$ is a matrix of dimension $n_1 \times n$ corresponding to the interaction coefficients of OTU $i \in G_1$. $A_2$ is a matrix of dimension $n_2 \times n$ rows corresponding to the interaction coefficients of OTU $i \in G_2$.
And  $f(s) \in \R^n$, such that 
\begin{align}
f_i(s) = \begin{cases}
\bar{\mu}_i \dfrac{s_1}{K_i + s_1} \quad \forall i \in G_1 \\
\bar{\mu}_i \dfrac{s_2}{K_i + s_2} \quad \forall i \in G_2
\end{cases}
\end{align}
With the above system \eqref{system}, is rewritten in equations \eqref{Eq1}, \eqref{Eq2}, \eqref{Eq3}, and \eqref{Eq4}.

\begin{align}
\label{Eq1}	\dot{x} = & \diag(x)(\diag(f(s))(1_{n\times 1} + Ax) - D_{n\times 1}) \\
\label{Eq2}	\dot{s_1} = & (s_{in}-s_1)D + \begin{bmatrix}
-(1_{n_1\times 1} +A_1x) \\0
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\
\label{Eq3}	\dot{s_2} = & -s_2D+
\begin{bmatrix}
1_{n_1\times 1} +A_1x \\ -(1_{n_2\times 1} +A_2x)
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\	
\label{Eq4}	\dot{s_3} = & -s_3D+\begin{bmatrix}
0	\\(1_{n_2\times 1} +A_2x) 
\end{bmatrix}^\top \diag(k) \diag(f(s))x 
\end{align}

@@

Note $J_I(x)$ the Jacobian matrix of function $\mathcal{I}$. then a first order approximation of $\I(\cdot)$ centred at $\bar{x}$ gives: $\I(x) = \I(\bar{x}) + J_I(\bar{x})(x-\bar{x}) + o(\Vert x- \bar{x} \Vert)$. When $\bar{x}= 0$ one recovers the growth expression from the previous section (equations \eqref{gLV growth1} and \eqref{gLV growth2}) implying that $J_I(\bar{0})$ can be seen as the interaction matrix from model \eqref{system}.

@@

The control $u(t)$ is intended to drive the system to be near a desired output $z(t)$, which in this context are the measurements of the concentrations of OTU. The term $\lambda \Vert(u -\vec{1}) \Vert_R$, was added for two reasons:

@@

The problem of approximating the solution of the system to a desired reference ($z$ in this case) is called the optimal tracking problem. For solving such a problem the approach developed by Cimen \textit{et al.} \cite{Cimen2004, Cimen2008} was adapted to our problem. The method proposed involves the resolution of Approximating Sequences of Ricatti Equations (ASRE).

@@

The tracking problem does not consider a constrained control. Nevertheless, the methods of Cimen \textit{et al.}\cite{Cimen2004} were directly used with an explicit constraint in the synthesis of the control. Even though this is probably suboptimal it worked fairly well.

Another departure from their method is that in their formulation an approximation of the dynamics for calculating the trajectories is used. They proved that such a linearisation converged to the original dynamics. In the case here presented the linearisation of the dynamics was not necessary.

@@

Finally for the parametric identification of a particular class of interaction functions an static optimization problem is proposed \eqref{Parameter Identification}. Suppose $u(t)$ is the optimal control of problem \eqref{Optimal Control}, and $x(t)$ is the associated optimal trajectory. Suppose a set of parameters $\theta \in \Theta$, such that the interaction function has a explicit form that depends on this parameters $I_{\theta}(x)$ ($\theta$ would represent matrix $A$ from section 2 for example). 
\begin{align}
\label{Parameter Identification} \min & \quad \Vert I_{\theta}(x(\cdot)) - u(\cdot) \Vert_{L^2([0,T])} \\
s.t. &  \quad \theta \in \Theta \notag \\
& \quad t \in [0,T] \notag
\end{align}

@@

For compensating the lack of a theoretical optimal control, the approach presented in here was first tested against synthetic data. The test consisted in simulating system \eqref{system} with growth functions \eqref{gLV growth1} and \eqref{gLV growth2} (therefore $I(x) = 1 + Ax$) for $n_1 = 10$ and $n_2 = 5$. In all figures, dotted lines represent the synthetic data, while the continuous lines represent the method's output.

The choice of matrices $Q$ and $R$ are always of discussion in a quadratic regulator. $Q$ represents the covariance matrix of measurements, while the interpretation of $R$ is not clear. In the simulations here presented $Q = I_n$ and $R = \begin{bmatrix}
10^{-4}I_{n_1} & 0 \\ 0 & 10^{-5}I_{n_2}
\end{bmatrix}$, found by trial and error in order for the control to fit accurately the species abundance, instead of the control approaching 1. The rational behind choosing a smaller value for the group $G_2$ is related to the one order of magnitude difference in yields between $G_1$ and $G_2$.

@@

The original interaction matrix and the fitted interaction matrix can be seen and compared in figure \eqref{Comparison Interaction Matrices}. It can be seen that the new matrix is very different from the original one, in both order of magnitude and sign. Implying that the model is probably not identifiable. The system was simulated again with the new matrix, and one can see that in general terms it captures the outcome of the system as shown in figures \ref{AOB LV simulation}, \ref{NOB LV simulation}, and \ref{Metabolites LV simulation}. The final and central message of this section is that if the interaction function is of the form $1+Ax$, one could be able to get a good fit from the data with the method here presented.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_AOB_Iter_5_LV}
	\caption{AOB trajectories using the identified matrix from figure \eqref{Comparison Interaction Matrices}.}
	\label{AOB LV simulation}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_NOB_Iter_5_LV}
	\caption{NOB trajectories using the identified matrix from figure \eqref{Comparison Interaction Matrices}.}
	\label{NOB LV simulation}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.5 \textwidth]{Synthetic_data_no_noise//221119_no_noise_iter_5_metabolites_Iter_5_LV}
	\caption{Metabolites trajectories using the identified matrix from figure \eqref{Comparison Interaction Matrices}.}
	\label{Metabolites LV simulation}
\end{figure}
